---
title: "Do Sea‑Level Rise Maps Backfire? A California‑Focused Causal Tutorial"
format: html
editor: visual
---

# Do Sea‑Level Rise Maps Backfire? A California‑Focused Causal Tutorial

## Background

Sea‑level rise (SLR) is one of California’s most pressing climate risks. Yet even in a climate‑aware state, public concern about SLR remains surprisingly low. Mildenberger et al. (2024) test whether personalized SLR maps, the kind used by governments and advocacy groups actually increase concern by showing residents how rising seas could affect their homes.

The surprising finding: Maps often reduce perceived personal harm, even for people whose homes are projected to flood.

This blog post walks through the design‑based causal method used in the study and recreates the California‑only version of Figure 4, which shows how the map’s treatment effect varies with distance from the SLR boundary.

# Study Design

## Randomized experiment

Respondents in California were recruited using address‑level mail‑to‑web sampling. Each respondent was randomly assigned to:

Treatment (treat_map = 1): See a customized SLR map centered on their home

Control (treat_map = 0): See no map

Because treatment is randomized, the difference in outcomes between treated and control respondents is a causal estimate of the map’s effect.

# Outcome variable

We focus on:

dv_personal — perceived personal harm from SLR (0–3 scale)

Running variable To recreate Figure 4, we use:

dist_min — distance (meters) from the SLR boundary

negative = inside projected flood zone

positive = outside projected flood zone

This running variable is used to estimate spatial heterogeneity in the treatment effect.

# Causal Identification Strategy

Randomized experiment is Average Treatment Effect (ATE) Because map exposure is randomly assigned, the causal effect is simply:

### (add the studies formula here)

Spatial heterogeneity (Figure 4 logic) To understand whether the map works differently for people closer to the flood zone, the authors:

Bin respondents into 50‑meter intervals of distance to the SLR boundary

Estimate the treatment effect within each bin

Plot those bin‑specific effects against distance

This is not a formal regression discontinuity design, but it borrows RD intuition:

If the map increases personal relevance, the treatment effect should be strongest near the boundary.

Instead, the effect is negative everywhere.

# Load and Clean California Data

```{r}
library(tidyverse)
library(broom)
library(estimatr)

# Load CA survey data
data_ca <- read_csv("data/SeaLevelRiseSurvey_California_January+16,+2021_13.54.csv") %>%
  .[-c(1,2),] %>%
  mutate(geo = "CA")

# Load CA distance-to-boundary file
dist_ca <- read_csv("data/ca_dist_mins.csv")

# Load CA combined metadata
cw_ca <- read_csv("data/ca_combined_final_withtraffic_sub.csv")

# Load flood zone file
fld_zone <- read.csv("data/fld_zone.csv")

```

# Merge Survey, Distance and Flood Zone

```{r}
dist_ca_clean <- dist_ca %>%
  filter(!duplicated(.)) %>%
  group_by(adds) %>%
  filter(dist_min == min(dist_min)) %>%
  inner_join(
    cw_ca %>% select(adds, email, FID) %>% filter(!duplicated(.)),
    by = "adds"
  ) %>%
  rename(RecipientEmail = email) %>%
  inner_join(
    fld_zone %>% select(adds, COUNTYFP, TRACTCE, FLD_ZONE) %>% filter(!duplicated(.)),
    by = "adds"
  )

```

# Construct California Only Analytic Dataset

```{r}
dv_recode <- function(x){
  case_when(
    x == "Not at all" ~ 0,
    x == "Only a little" ~ 1,
    x == "A moderate amount" ~ 2,
    x == "A great deal" ~ 3
  )
}

data_ca_clean <- data_ca %>%
  left_join(dist_ca_clean, by = "RecipientEmail") %>%
  mutate(
    treat_map = !is.na(exp_emphasis),
    slr = str_detect(img, "SLR"),
    dv_personal = dv_recode(slharmpersonal),
    dist_min = ifelse(slr == TRUE, -dist_min, dist_min)
  ) %>%
  filter(!is.na(dv_personal), !is.na(dist_min))

```

# Step 1 - Estimate the ATE of California

```{r}
mod_ate <- lm_robust(dv_personal ~ treat_map, se_type = "HC1", data = data_ca_clean)
summary(mod_ate)

```

# Step 2 — Recreate Figure 4

```{r}
bin_width <- 50

data_ca_binned <- data_ca_clean %>%
  mutate(
    dist_bin = bin_width * round(dist_min / bin_width),
    dist_bin = pmax(pmin(dist_bin, 500), -500)
  )

```

# Step 3 - Estimate treatment effect within each bin

```{r}
effects_ca <- data_ca_binned %>%
  group_by(dist_bin) %>%
  filter(n() >= 15) %>%
  do(tidy(lm_robust(
    dv_personal ~ treat_map,
    se_type = "HC1",
    data = .
  ))) %>%
  ungroup() %>%
  filter(term == "treat_mapTRUE") %>%
  mutate(
    ci_low = estimate - 1.96 * std.error,
    ci_high = estimate + 1.96 * std.error
  )

```

# Step 4 - Plot California- Only Figure

```{r}
ggplot(effects_ca, aes(x = dist_bin, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 20, alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  theme_minimal() +
  labs(
    title = "Outcome: Sense of personal harm (California only)",
    subtitle = "Treatment effect of map by distance from SLR boundary",
    x = "Distance from SLR boundary (negative = inside SLR zone)",
    y = "Treatment effect of map"
  )

```

# Interpretation

Three key insights emerge from the California analysis:

Maps reduce perceived personal harm across almost all distances.\
Even respondents whose homes are projected to flood show lower concern after seeing the map.

There is no sharp change at the SLR boundary.\
If maps increased personal relevance, we would expect a spike near 0 meters. Instead, the curve is flat and negative.

California behaves like the national sample.\
Despite higher climate awareness, the backfire effect persists.

# Conclusion

This study is a powerful example of design‑based causal inference:

Randomization leads to unbiased treatment effect

Spatial heterogeneity relates to mechanism exploration

Simple models allow for transparent interpretation

For policymakers, the takeaway is clear: SLR maps may unintentionally reduce concern, even among at‑risk households.
